---
title: 'Looking Both Ways - Infix Functions'
author: "Matt Cole"
subtitle: "Some %R% tips"
date: 2017-02-05
output: md_document
categories: ["R-package"]
tags: ["R", "package"]
---

In your R journeys you may have come across some interesting functions like `apply` statements or even `lm`. One function that is particularly helpful (and interesting) is the piping operator (`%>%`) from the [magrittr pacakge](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html). You may have noticed that the piping operator is similar to the matrix multiplcation operator `%*%`, in that they are both _sandwitch_ functions (may or may not be trying to coin this term right now), as the function call is/are a symbol(s) enclosed by a `%` on both times. These sandwitch functions in R are actually members of a larger class of functions, known as  _infix_ [functions](https://en.wikipedia.org/wiki/Infix_notation). Unlike most functions such as `mean()`, `summary()`, or `kable()`, are _prefix_ functions, which take their arguments after the fucntion is called (`mean(c(1.2,1.6,0.4,3.1)`). Infix fuctions on the other hand, come inbetween its (two) arguments. Other infix functions include basic addition, and subtraction (`+`, `-`) and all your other common aresthmatic functions. Many in R however, are functions enclosed by a % on both side to indicate their special features. Some other examples are `%*%` (matrix multiplication), or `%in%`, [etc](http://adv-r.had.co.nz/Functions.html).

For example:

```{r}
matrix(c(1:4),2) %*% matrix(c(1,0,0,1),2)
```

We can even define our own infix functions as follows:

```{r}
`%+2%` <- function(x, y){
return(x + y + 2)
}
```

So, what would `1 %+2% 1` result in?

In short, while these functions are, deep down, just regular functions. They can improve readability considerably in your code - imagine needing to use `add(x,y)` whenever you had to find the sum of two numbers.


`(2+2) * 10 - 6` would turn into `subtract(multiply(add(2,2),10),6)`. What a monster!
